---
title: "Data preprocessing"

project:
  execute-dir: project

format:
  html:
    theme:
      - cosmo
      - brand
    css: styles.css
    toc: true

execute:
  freeze: auto
---

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(gfwr)
library(raster)
library(sf)
library(janitor)
library(countrycode)
library(openxlsx)
library(forcats)
library(stringr)
library(worrms)
library(purrr)
```

This notebook details the data sources used in our analysis, and performs data preparation processing steps.

The following hyperparameters allow the user to control the downloading of external files and rewriting of local files:

-   `FORCE_WRITE == FALSE` means that local files (outputs of this notebook) will not be rewritten if they already exist. They will be if the parameter is set to `TRUE`.
-   `FORCE_DOWNLOAD_GFW` controls in a similar manner the re-downloading of Global Fishing Watch (GFW) data from the GFW API.
-   `SKIP_CATCH` lets the notebook ignore the catch data processing, as it is time-consuming and independent from the rest.

```{r}
FORCE_WRITE <- FALSE
FORCE_DOWNLOAD_GFW <- FALSE
SKIP_CATCH <- TRUE
```

## Reference polygons

Several geo-referenced polygons are used:

-   The shape of the FAO major fishing area 34 (Atlantic, Eastern-Central)

```{r}
#| label: load-fao34-sf

fao34_sf <- st_read("data/shapefiles/FAO_AREAS_CWP_NOCOASTLINE/FAO_AREAS_CWP_NOCOASTLINE.shp") |> filter(F_CODE == 34)
```

-   Exclusive Economic Zones (EEZs) intersecting the FAO 34 area, and their intersection with the latter.

```{r}
#| label: load-eez-sf

eez_sf <- st_read("data/shapefiles/FAO34_EEZ/eez_fao34.shp")
eez_overlap_sf <- eez_sf |>
  st_intersection(fao34_sf |> dplyr::select(geometry))
```

-   Large Marine Ecosystems (LMEs) of the Canary Current and the Gulf of Guinea.

```{r}
#| label: load-lme-sf

lme_sf <- st_read("data/shapefiles/lme-large-marine-ecosystems-main/polygons/shp/lme66.shp") |>
  clean_names() |>  
  filter(lme_number %in% c(27, 28))
```

## Rasters

### Bathymetry raster

Global altitude raster in meters, available on GFW's data portal ([link](https://globalfishingwatch.org/data-download/datasets/public-bathymetry-v1)). Source: General Bathymetric Chart of the Oceans (GEBCO).

```{r}
#| label: load-bathymetry-raster

bathymetry_raster <- raster("data/bathymetry.tif")
crs(bathymetry_raster) <- CRS("+proj=longlat +datum=WGS84 +no_defs")
```

### Distance to shore raster

Global raster indicating distance from the closest shore for any point in the ocean. Available on GFW's data portal ([link](https://globalfishingwatch.org/data-download/datasets/public-distance-from-shore-v1)). Units: kilometers. Resolution: 1km. Source: *Pacific Islands Ocean Observing System's Distance to Nearest Coastline: 0.01-Degree Grid: Ocean dataset*.

```{r}
#| label: load-dist-to-shore-raster

distance_to_shore_raster <- raster("data/distance-from-shore.tif")
crs(distance_to_shore_raster) <- CRS("+proj=longlat +datum=WGS84 +no_defs")
```

### Distance from port raster

In the context of industrial fisheries, this raster provides the distance from the closest port for any point in the ocean. This dataset is used to remove potentially erroneous fishing hours close to fishing ports. Available on GFW's data portal ([link](https://globalfishingwatch.org/data-download/datasets/public-distance-from-port-v1)). Units: meters. Resolution: 1 km. Source: The port distances are calculated using the [Global Fishing Watch Anchorages dataset](https://globalfishingwatch.org/datasets-and-code/anchorages/).

```{r}
#| label: load-dist-from-port-raster

distance_from_port_raster <- raster("data/distance-from-port-v20201104.tiff")
crs(distance_from_port_raster) <- CRS("+proj=longlat +datum=WGS84 +no_defs")
```

## Vessel characteristics of the global apparent fishing fleet

Length over all (LOA, in meters), maximum engine power (in kW) and gross tonnage (GT, in GT), either drawn from vessel registers or estimated using GFW's fishing tracks neural network, for each MMSI number in the global apparent fishing fleet from 2012 to 2024. Primary key is the (year x MMSI) pair. [Link](https://globalfishingwatch.org/data-download/datasets/public-fishing-effort).

```{r}
#| label: load-vessels-info

vessels_info <- read_csv("data/vessel_info_gfw/fishing-vessels-v3.csv")
```

## Apparent fishing effort

### Downloading the fishing effort within the FAO 34 area

Apparent fishing effort data spanning a 5-year period (2020-01-01 to 2024-12-31) were downloaded using the *gfwr* R package. We used the `get_raster` function to import the fishing hours of cells contained within the FOA 34 area polygon. The GFW identifier in this dataset is the `vessel_id` variable, produced by the aggregation of track belonging to the same vessel using conservative criteria.

```{r}
#| label: gfw-import-fun

# Import function embedding get_raster
gfw_import <- function(year, spatial_resolution, temporal_resolution, region, region_source){
  get_raster(spatial_resolution = spatial_resolution,
             temporal_resolution = temporal_resolution,
             group_by = "VESSEL_ID",
             start_date = paste0(year,"-01-01"),
             end_date = paste0(year,"-12-31"),
             region = region,
             region_source = region_source,
             key = Sys.getenv("GFW_TOKEN"))
}
```

Fishing effort data are downloaded with at separate spatio-temporal resolutions:

1.  Low spatial resolution (0.1°x0.1°) and monthly temporal resolution.
2.  High spatial resolution (0.01°x0.01°) with a yearly temporal resolution.

```{r}
#| label: gfw-monthly-low-import-fun

temp_res <- "MONTHLY"
spatial_res <- "LOW"
years <- 2020:2024

raw_file_path <- paste0("output/data/gfw_apparent_fishing_effort/gfw_effort_FAO34_", temp_res, "_", spatial_res, "_", min(years), "-", max(years), "_RAW.csv")

if (!file.exists(raw_file_path) | FORCE_DOWNLOAD_GFW) {
  gfw_data_ML <- lapply(
    years, 
    function(year)
      gfw_import(
        year = year, 
        spatial_resolution = spatial_res,
        temporal_resolution = temp_res,
        region = fao34_sf, 
        region_source = 'USER_SHAPEFILE'
      )
  ) |> 
    bind_rows() |> 
    clean_names()

  write.csv(gfw_data_ML,
            raw_file_path,
            row.names = FALSE)
} else {
  gfw_data_ML <- read_csv(raw_file_path)
}
```

```{r}
#| label: gfw-yearly-high-import-fun

temp_res <- "YEARLY"
spatial_res <- "HIGH"
years <- 2020:2024

raw_file_path <- paste0("output/data/gfw_apparent_fishing_effort/gfw_effort_FAO34_", temp_res, "_", spatial_res, "_", min(years), "-", max(years), "_RAW.csv")

if (!file.exists(raw_file_path) | FORCE_DOWNLOAD_GFW) {
  gfw_data_YH <- lapply(
    years, 
    function(year) {
      message("\nProcessing year: ", year)
      tryCatch({
        gfw_import(
          year = year, 
          spatial_resolution = spatial_res,
          temporal_resolution = temp_res,
          region = fao34_sf, 
          region_source = 'USER_SHAPEFILE'
        )}, 
        error = function(e) {
         message("Failed to retrieve data for year ", year, ": ", e$message)
         NULL  # Skip this year
        })
    }
  ) |> 
    bind_rows() |> 
    clean_names()

  write.csv(gfw_data_YH,
            raw_file_path,
            row.names = FALSE)
} else {
  gfw_data_YH <- read_csv(raw_file_path)
}
```

### First cleaning

```{r}
#| label: gfw-data-first-cleaning

first_cleaning_gfw_data <- function(gfw_data) {
  gfw_data |> 
    filter(vessel_type == "FISHING") |>                    # Filtering out non "FISHING" vessels
    mutate(year = as.numeric(str_sub(time_range, 1, 4))) |>  # Extract year from time_range
    dplyr::select(-c(entry_timestamp, exit_timestamp, first_transmission_date, last_transmission_date, vessel_type)) 
}

gfw_data_ML <- gfw_data_ML |> first_cleaning_gfw_data()

gfw_data_YH <- gfw_data_YH |> first_cleaning_gfw_data()
```

### Spatial joins

Effort data is joined spatially several times. First with the EEZ polygons in order to allocate fishing hours to the waters of each coastal state. Then with the raster.

```{r}
#| label: gfw-data-spatial-join

spatial_join_gfw_data <- function(gfw_data) {
  
  gfw_data <- gfw_data |> 
    st_as_sf(coords = c("lon", "lat"), crs = st_crs(eez_sf), remove = FALSE) |>  # Convert spatial effort into and sf object with CRS = WGS 84
    st_join(eez_sf |>                                                            # Spatial join with EEZs
            filter(POL_TYPE != "Joint regime") |>
            dplyr::select(TERRITORY1, ISO_TER1, geometry), 
          join = st_intersects, left = TRUE) |> 
    rename(eez_name = TERRITORY1, eez_iso3c = ISO_TER1) |> 
    mutate(
      eez_iso3c = case_when(
        eez_name == "Madeira" ~ "PRT",
        eez_name == "Azores" ~ "PRT",
        eez_name == "Canary Islands" ~ "ESP",
        TRUE ~ eez_iso3c
      )
    )
  
  # Extract values from rasters
  gfw_data$distance_to_shore_km <- raster::extract(distance_to_shore_raster, gfw_data)
  gfw_data$distance_from_port_km <- raster::extract(distance_from_port_raster, gfw_data)
  gfw_data$bathymetry_m <- raster::extract(bathymetry_raster, gfw_data)
  
  return(gfw_data |> st_drop_geometry())
}

joined_file_path <- paste0("output/data/gfw_apparent_fishing_effort/gfw_effort_FAO34_MONTHLY_LOW_2020-2024_JOINED.csv")
if (!file.exists(joined_file_path) | FORCE_WRITE) {
  gfw_data_ML <- spatial_join_gfw_data(gfw_data_ML)
  write.csv(gfw_data_ML,
            joined_file_path,
            row.names = FALSE)
} else {
  gfw_data_ML <- read_csv(joined_file_path)
}

joined_file_path <- paste0("output/data/gfw_apparent_fishing_effort/gfw_effort_FAO34_YEARLY_HIGH_2020-2024_JOINED.csv")
if (!file.exists(joined_file_path) | FORCE_WRITE) {
  gfw_data_YH <- spatial_join_gfw_data(gfw_data_YH)
  write.csv(gfw_data_YH,
            joined_file_path,
            row.names = FALSE)
} else {
  gfw_data_YH <- read_csv(joined_file_path)
}
```

The "YEARLY HIGH" table will only be used for the study of fishing within territorial water. We then filter the effort that occured less than 200 nm from shore.

```{r}
gfw_data_YH <- gfw_data_YH |> 
  filter(distance_to_shore_km <= 200 * 1.852)
```

### Joining with vessel characteristics

The characteristics (size, power, and tonnage) of the vessel are added to each fishing observation. The information dataset is also filtered to retain only those vessels present in the analyzed effort data.

```{r}
#| label: gfw-data-add-vessels-info

vessels_info <- vessels_info  |> 
  mutate(
    is_inferred_length = is.na(length_m_registry),
    is_inferred_power = is.na(engine_power_kw_registry),
    is_inferred_tonnage = is.na(tonnage_gt_registry)
  ) |> 
  dplyr::select(year, mmsi, length_m_gfw, engine_power_kw_gfw, tonnage_gt_gfw, is_inferred_length, is_inferred_power, is_inferred_tonnage, registries_listed) |>
  distinct()

add_vessel_characteristics <- function(gfw_data) {
  gfw_data |> 
  left_join(vessels_info |> 
              distinct(year, mmsi, length_m_gfw, engine_power_kw_gfw, tonnage_gt_gfw),
            by = c("year", "mmsi"),
            relationship = "many-to-one")
}

gfw_data_ML <- gfw_data_ML |> 
  add_vessel_characteristics() 

gfw_data_YH <- gfw_data_YH |> 
  add_vessel_characteristics()

all_vessels_x_years <- gfw_data_ML |> 
  distinct(year, mmsi, vessel_name)

vessels_info <- all_vessels_x_years |>
  left_join(vessels_info,
            by = c("year", "mmsi"),
            relationship = "many-to-one") |>
  relocate(vessel_name, .after = mmsi)
```

The apparent effort in kW.hours can then be calculated as the product of the maximum engine power and the apparent fishing hours:

```{r}

gfw_data_ML <- gfw_data_ML |> 
  mutate(kWh = engine_power_kw_gfw * apparent_fishing_hours)

gfw_data_YH <- gfw_data_YH |> 
  mutate(kWh = engine_power_kw_gfw * apparent_fishing_hours)
```

### Cleaning

Several variables are added/modified. The `gear_type` label indicating the type of gear is simplified. The 18-class hierarchical classification provided by Global Fishing Watch is summarized into 8 classes:

| Original labels `gear_type` | New labels `gear_type` | French translation `gear_type_fr` |
|:----------------------:|------------------------|------------------------|
| Fishing OR Other gear OR Inconclusive | Unknown | Inconnus |
| Fishing / Seiners | Other seines | Autres sennes |
| Fishing / Seiners / Other seines | Other seines | Autres sennes |
| Fishing / Seiners / Purse seines | Other seines | Autres sennes |
| Fishing / Seiners / Purse seines / Tuna purse seines | Tuna purse seines | Sennes coulissantes au thon |
| Fishing / Seiners / Purse seines / Other purse seines | Other seines | Autres sennes |
| Fishing / Squid jiggers | Squid jiggers | Turluttes |
| Fishing / Drifting longlines | Hooks and lines | Lignes et hameçons |
| Fishing / Pole and line | Hooks and lines | Lignes et hameçons |
| Fishing / Trollers | Hooks and lines | Lignes et hameçons |
| Fishing / Fixed gear | Fixed gear | Arts dormants |
| Fishing / Fixed gear / Pots and traps | Fixed gear | Arts dormants |
| Fishing / Fixed gear / Set longlines | Fixed gear | Arts dormants |
| Fishing / Fixed gear / Set gillnets | Fixed gear | Arts dormants |
| Fishing / Trawlers | Trawls | Chaluts |
| Fishing / dredgers | Dredges | Dragues |

```{r}
#| label: gfw-data-clean-gear-types-fun

clean_gear_types <- function(gfw_data) { 
  gfw_data |> 
    mutate(
      gear_type = case_when(
        grepl("TUNA", gear_type) ~ "Tuna purse seines",
        grepl("SEINE", gear_type) & !grepl("TUNA", gear_type) ~ "Other seines",
        grepl("SQUID", gear_type) ~ "Squid jiggers",
        grepl("TRAWL", gear_type) ~ "Trawls",
        grepl("DREDGE", gear_type) ~ "Dredges",
        grepl("FIXED|POT|SET", gear_type) ~ "Fixed gear",
        grepl("LINE|TROLL", gear_type) ~ "Hooks and lines",
        grepl("FISHING|OTHER|GEAR|INCONCLUSIVE", gear_type) ~ "Unknown",
        TRUE ~ gear_type
      ),
      gear_type_fr = case_when(
        gear_type == "Tuna purse seines" ~ "Sennes coulissantes au thon",
        gear_type == "Other seines" ~ "Autres sennes",
        gear_type == "Squid jiggers" ~ "Turluttes",
        gear_type == "Trawls" ~ "Chaluts",
        gear_type == "Dredges" ~ "Dragues",
        gear_type == "Fixed gear" ~ "Arts dormants",
        gear_type == "Hooks and lines" ~ "Lignes et hameçons",
        gear_type == "Unknown" ~ "Inconnus",
        TRUE ~ gear_type
      )
    )
}
```

Labels are also created to characterize the geopolitical affiliation of flags and fishing zones. This distinguishes between EEZs on the high seas and flags belonging to the European Union, a coastal nation in FAO zone 34, or “West Africa,” defined as the African countries in FAO area 34.

```{r}
#| label: gfw-data-add-geopo-labels-fun

west_africa_flags <- c("MAR", "ESH", "MRT", "SEN", "GMB", "GNB",
                       "GIN", "SLE", "LBR", "CPV", "CIV", "GHA",
                       "TGO", "BEN", "NGA", "CMR", "STP", "GNQ",
                       "GAB", "COD", "COG", "AGO")

add_political_labels <- function(gfw_data) {
  gfw_data |> 
    mutate(
      
      # Add a region type label to distinguish effort in EEZs vs. not
      region_type = ifelse(!is.na(eez_name), "EEZ", "High seas"),
      region_type_fr = ifelse(region_type == "EEZ", "ZEE", "Haute mer"),
      
      # Clean the flag labels
      flag_iso3c = flag,
      flag = countrycode(flag, origin =  "iso3c", destination = "country.name"),
      
      # Add booleans for flag political or geographical belonging
      #is_european_flag = ifelse(flag_iso3c %in% pull(read.xlsx("https://www.dropbox.com/s/dp8l4nz8fmr5v10/eu27_iso_codes.xlsx?dl=1")), TRUE, FALSE),
      is_fao34_flag = flag_iso3c %in% unique(eez_overlap_sf$ISO_TER1),
      is_west_african_flag = flag_iso3c %in% west_africa_flags,
      
      # Add booleans for type of fishing (domestic or distant)
      domestic = (region_type == "EEZ") & (flag_iso3c == eez_iso3c) | (eez_iso3c == "ESH") & (flag_iso3c == "MAR"),  # Moroccan fishery in West. Sahara is domestic here
      
      ## Type of origin of the flag
      origin_type_fr = case_when(
        is.na(flag_iso3c) ~ NA,
        domestic ~ "Domestique",
        is_west_african_flag ~ "Distante (pavillon ouest-africain)",
        TRUE ~ "Distante (autre)"
      ),
      origin_type_fr = factor(origin_type_fr, levels = c("Domestique", "Distante (pavillon ouest-africain)", "Distante (autre)")
      ),
      origin_type_fr = fct_na_value_to_level(origin_type_fr),
      
      ## Type of fishing ground
      destination_type_fr = case_when(
        region_type == "High seas" ~ "Haute mer",
        domestic ~ "Domestique",
        region_type == "EEZ" & !domestic ~ "ZEE étrangère"
      )
    )
}
```

Validity of the IMO numbers provided by Global Fishing Watch is checked. In addition to being composed of seven digits, IMO numbers for vessels use a checksum system. The digits of the six highest powers of 10 (2 to 6) are added together, multiplied each by their power of 10, and the units digit of the result must be the units digit of the IMO number. Example: 8820274. We have 8x7 + 8x6 + 2x5 + 0x4 + 2x3 + 7x2 = 134. 4 is also the units digit of 8820274. This number is valid.

```{r}
#| label: gfw-data-imo-checksum-fun

checksum_imo <- function(numbers) {
  numbers <- as.character(numbers)
  
  # Filter valid 7-digit numeric strings
  valid <- grepl("^\\d{7}$", numbers)

  # Compute checksum and compare to check digit
  result <- rep(NA_character_, length(numbers))
  
  if (any(valid)) {
    valid_numbers <- numbers[valid]
    
    digits <- matrix(as.integer(strsplit(valid_numbers, "") |> unlist()), ncol = 7, byrow = TRUE)
    
    weights <- 2:7
    weighted_sum <- rowSums(digits[, 1:6] * weights)
    check_digit <- digits[, 7]
    
    valid_imo <- check_digit == (weighted_sum %% 10)
    
    result[valid] <- ifelse(valid_imo, valid_numbers, NA)
  }
  
  return(result)
}
```

Here, we clean the types of vessel, add labels relating to the states and type of fishing (distant water or domestic), extract the year from `time_range`, and check the IMO numbers.

```{r}
#| label: gfw-data-clean-labels

clean_labels_gfw_data <- function(gfw_data) {
  gfw_data |> 
    clean_gear_types() |> 
    add_political_labels() |> 
    mutate(
      imo = checksum_imo(imo)                         # Check that IMOs are real IMOs
    ) 
}

gfw_data_ML <- gfw_data_ML |> clean_labels_gfw_data()
gfw_data_YH <- gfw_data_YH |> clean_labels_gfw_data()
```

Columns are rearranged.

```{r}
#| label: gfw-data-relocate-columns

relocate_gfw_data <- function(gfw_data) {
  gfw_data |> 
    relocate(
      year, time_range,
      lat, lon,
      region_type, region_type_fr,
      eez_name, eez_iso3c,
      flag, flag_iso3c,
      vessel_name, mmsi, imo, call_sign,
      apparent_fishing_hours, kWh,
      gear_type, gear_type_fr,
      #is_european_flag, 
      is_fao34_flag, is_west_african_flag,
      domestic, origin_type_fr, destination_type_fr,
      length_m_gfw, engine_power_kw_gfw, tonnage_gt_gfw,
      distance_to_shore_km, distance_from_port_km, bathymetry_m,
      vessel_id
    )
}

gfw_data_ML <- gfw_data_ML |> relocate_gfw_data()
gfw_data_YH <- gfw_data_YH |> relocate_gfw_data()
```

Finally, the data is filtered. Only vessels classified as “FISHING” are retained, in order to exclude the apparent effort of vessels that are not considered fishing vessels by Global Fishing Watch. The columns relating to the vessel's entry and exit from the cell, as well as those concerning its AIS transmission period, are removed.

Above all, limits are applied to the distance between the fishing event and the closest port (5 km) and the minimum number of fishing hours for a vessel (2,000 hours in 5 years). Individual fishing time can be calculated by Global Fishing Watch identifier (`vessel_id`), MMSI number, or IMO number. The first case tends to underestimate fishing time since the same MMSI number can be associated with several `vessel_id`s. The last case is ideal, but data on IMO numbers is very incomplete. We therefore choose the MMSI number.

```{r}
clean_gfw_data <- function(gfw_data,                                 # GFW apparent fishing effort data after spatial joins
                           min_dist = 5,                             # Minimum distance from port in km
                           min_hours = 2000,                         # Minimum apparent fishing time per vessel (vessel below this threshold are removed)
                           unit = c("mmsi", "vessel_id", "imo"),     # Vessel identifier to use
                           save_small_efforts = FALSE) {             # Whether to include the effort of vessel contributing below the minimum time treshold in a separate df
  
  # Match the 'unit' argument
  unit = match.arg(unit)
  unit_column <- switch(unit,
                        "mmsi" = gfw_data$mmsi,
                        "vessel_id" = gfw_data$vessel_id,
                        "imo" = gfw_data$imo)
  gfw_data <- gfw_data |> mutate(unit = unit_column)
  
  # Save raw effort
  total_effort_raw <- sum(gfw_data$apparent_fishing_hours, na.rm = TRUE)
  
  # First filters
  gfw_data <- gfw_data |> 
    filter(distance_from_port_km >= min_dist)        # Distance from port threshold
  
  # Selecting the vessels to keep / filter out based on individual effort
  total_effort <- sum(gfw_data$apparent_fishing_hours, na.rm = TRUE)
  
  vessel_list <- gfw_data |> 
    summarize(apparent_fishing_hours = sum(apparent_fishing_hours, na.rm = TRUE),
              .by = c(unit)) |> 
    filter(apparent_fishing_hours < min_hours) |> 
    pull(unit)
  
  discarded_effort <- gfw_data |> 
    filter(unit %in% vessel_list) |>
    summarize(total = sum(apparent_fishing_hours, na.rm = TRUE)) |> 
    pull(total)
  
  # (Optional) Saving the effort of vessels fishing less than min_hours
    if (save_small_efforts) {
    small_efforts <- gfw_data |>  filter(unit %in% vessel_list) |> dplyr::select(-unit)
  } else {
    small_efforts <- NULL
  }
  
  # Filtering out vessels below the threshold
  gfw_data <- gfw_data |> filter(!(unit %in% vessel_list)) |> dplyr::select(-unit)

  list(
    "initial_fishing_time" = total_effort_raw,
    "total_fishing_time" = total_effort, 
    "small_efforts_fishing_time" = discarded_effort,
    "clean_effort" = gfw_data,
    "small_efforts" = small_efforts
    )
}

gfw_ds_ML <- clean_gfw_data(gfw_data_ML, 
                            min_dist = 5, 
                            min_hours = 2000,
                            unit = "mmsi",
                            save_small_efforts = TRUE)

gfw_ds_YH <- clean_gfw_data(gfw_data_YH,
                            min_dist = 5,
                            min_hours = 2000,
                            unit = "mmsi",
                            save_small_efforts = FALSE)
```

The two datasets are save as .rds files:

```{r}
#| label: save-gfw-effort-ds

# Saving gfw effort datasets
if (!file.exists("output/clean_data/_temp/gfw_effort_FAO34_ML_0.rds") | FORCE_WRITE) {
  saveRDS(gfw_ds_ML, "output/clean_data/_temp/gfw_effort_FAO34_ML_0.rds")
}
if (!file.exists("output/clean_data/_temp/gfw_effort_FAO34_YH_0.rds") | FORCE_WRITE) {
  saveRDS(gfw_ds_YH, "output/clean_data/_temp/gfw_effort_FAO34_YH_0.rds")
}

# Saving vessel information dataset
if (!file.exists("output/clean_data/vessels_info.rds") | FORCE_WRITE) {
  saveRDS(vessels_info, "output/clean_data/vessels_info.rds")
}
```

```{r}
#| label: remove-all-reload-gfw-ds

rm(list = setdiff(ls()[grepl("gfw", ls())], lsf.str()))
effort <- readRDS("output/clean_data/_temp/gfw_effort_FAO34_ML_0.rds")$clean_effort
```

## Ownership

### Generating a list of vessels for identity and ownership investigation

We use the apparent fishing effort dataset ("MONTHLY LOW") to rank `vessel_id`'s by decreasing number of fishing hours (in the FAO 34 area between 2020 and 2024). We then save this ranking, including all the available identity information (GFW identifier, MMSI, IMO, vessel name, call sign, flag).

```{r}
#| label: ownership-vessel-list

vessels <- effort |> 
  summarize(apparent_fishing_hours = sum(apparent_fishing_hours, na.rm = TRUE),
            .by = c(flag, vessel_id, mmsi, vessel_name, imo, call_sign)) |> 
  arrange(desc(apparent_fishing_hours)) |> 
  mutate(
    effort_cumulated_percent = 100 * cumsum(apparent_fishing_hours) / sum(apparent_fishing_hours),
    rank = row_number()
  )

# Save for 

write.xlsx(vessels,
           "output/data/ownership/vessel_ranking_for_ownership_FAO34.xlsx")
```

### Manual identity investigation

This list is then manually supplemented with information from the [GISIS: Ship and Company Particulars](https://gisis.imo.org/Public/SHIPS/Default.aspx) database. The goal is to assign an IMO numbers (and if available a direct owner company) to each `vessel_id`.

Part of the research also involves comparing the AIS tracks of `vessel_id`'s suspected of being the same vessel. This comparison is carried out using Global Fishing Watch's [Vessel Viewer](https://globalfishingwatch.org/map/vessel-search). Sites such as [MarineTraffic](https://www.marinetraffic.com) are used to investigate the identity of vessels.

Accuracy requires the use of `vessel_id`s as identifiers, rather than MMSI numbers. This slightly increases the number of individuals to be identified. Identity searches are conducted on GISIS for all `vessel_id`s with more than 2,000 hours of apparent fishing between 2020 and 2024 in FAO area 34. Since several `vessel_id`s can be linked to a single MMSI number, some of the vessels retained in the `gfw_ds_ML$clean_effort` effort dataset have not been examined. However, it is expected that the vessels not examined be rather small contributors to the total apparent fishing effort.

### IMO number / sirect owner lists for ORBIS query

The supplemented ranking is then treated to create two lists: one of IMO numbers and one of GISIS-based direct owners. Those list will be used as input for two separate ORBIS queries.

```{r}
path <- "data/ownership/vessel_ranking_for_ownership_FAO34_completed.xlsx"

imo_completed_ranking <- read.xlsx(xlsxFile = path,
                                   sheet = "complete_imo")


# Select only vessel ids with 2000+hrs and clean --------------------------

imo_completed_ranking <- imo_completed_ranking |> filter(apparent_fishing_hours >= 2000)

imo_completed_ranking <- imo_completed_ranking |> 
  dplyr::select(-gfw_imo) |> 
  #rename(vessel_name = gisis_vessel_name) |> 
  mutate(
    company_address_country = case_when(
      is.na(company_address_country) ~ company_nationality,   # Fill company address
      TRUE ~ company_address_country
    ),
    imo = as.numeric(imo),                                    # Numeric type for IMO
    effective_since_year = as.numeric(effective_since_year),  # Numeric type for year
    gisis_vessel_name = str_trim(gisis_vessel_name),          # Trim white spaces to avoid false differences
    registered_owner = str_trim(registered_owner)
  ) |> 
  mutate(
    imo_confidence_status = case_when(
      imo_confidence_code == 1 ~ "IMO provided by GFW",       # Explicit the confidence code
      imo_confidence_code == 2 ~ "Direct match in GISIS",
      imo_confidence_code == 3 ~ "Non-ambiguous indirect match",
      imo_confidence_code == 4 ~ "Ambiguous indirect match",
      imo_confidence_code == 5 ~ "No association"
    )
  )


# Create summary of IMO ownership -----------------------------------------

imo_ownership <- imo_completed_ranking |>
  dplyr::select(vessel_id, mmsi, vessel_name, gisis_vessel_name, imo_confidence_code, imo_confidence_status, imo, registered_owner, company_imo, company_nationality, company_address_country, effective_since_year) |> 
  filter(!is.na(imo)) |>
  distinct() |> 
  arrange(company_nationality, company_imo, imo)

# Get IMO list for ORBIS search -------------------------------------------

imo_list <- imo_completed_ranking |> 
  dplyr::select(imo, gisis_vessel_name) |> 
  filter(!is.na(imo)) |>
  distinct()


# Get companies list for ORBIS search -------------------------------------

owners_list <- imo_completed_ranking |> 
  summarize(n_vessels = n_distinct(imo),
            .by = c(registered_owner, company_imo, company_nationality, company_address_country)) |> 
  arrange(registered_owner)

# Save in new sheets ------------------------------------------------------

wb <- loadWorkbook("data/ownership/vessel_ranking_for_ownership_FAO34_completed.xlsx")
addWorksheet(wb,"GFW-GISIS_key")
addWorksheet(wb,"IMO_list")
addWorksheet(wb, "owners_list")
writeData(wb, "GFW-GISIS_key", imo_ownership)
writeData(wb, "IMO_list", imo_list)
writeData(wb, "owners_list", owners_list)
saveWorkbook(wb, "output/data/ownership/vessel_ranking_for_ownership_FAO34_completed_processed.xlsx", overwrite = TRUE)
```

### Cleaning

Both lists (IMO numbers and names of owner companies) are used as queries in [Bureau Van Dijk's ORBIS](https://promo.globaldatabase.com/bvd/?utm_term=orbis%20bureau%20van%20dijk&utm_campaign=Bureau+van+Dijk&utm_source=adwords&utm_medium=ppc&hsa_acc=1811458963&hsa_cam=21662561486&hsa_grp=167632491515&hsa_ad=712026503454&hsa_src=g&hsa_tgt=kwd-355591263295&hsa_kw=orbis%20bureau%20van%20dijk&hsa_mt=p&hsa_net=adwords&hsa_ver=3&gad_source=1&gad_campaignid=21662561486&gclid=CjwKCAjw6vHHBhBwEiwAq4zvAzAGsaBVIIBd5PpPhVu2bRp-Lfi9hUx4YTSxdNsizwo-9fBB2cJ2yBoCTb4QAvD_BwE) database. The results are pre-processed manually. The data is then transformed manually to obtain a harmonized ownership database. Priority is given to information obtained via ORBIS, with information obtained from GISIS used as a secondary source.

```{r}

# Key for matching a GFW vessel (vessel_id) to the vessel IMO number and the GISIS company IMO number
gfw_gisis_key <- read.xlsx("output/data/ownership/vessel_ranking_for_ownership_FAO34_completed_processed.xlsx", sheet = "GFW-GISIS_key") |> clean_names()

# Result of the ORBIS search based on IMO numbers
orbis_vessels <- read.xlsx("data/ownership/export_orbis_FAO34.xlsx", sheet = "clean_imo_results") |> clean_names()

# Result of the ORBIS search based on GISIS owner company names
orbis_companies <- read.xlsx("data/ownership/export_orbis_FAO34.xlsx", sheet = "clean_gisis_owner_results") |> clean_names()


# Join

# First we add ownership information for the vessel which were recognized by ORBIS using their IMO number
ownership_table_1 <- gfw_gisis_key |> 
  dplyr::select(vessel_id, mmsi, vessel_name, imo_confidence_code, imo_confidence_status, imo) |> 
  left_join(orbis_vessels |> 
              dplyr::select(starts_with(c("imo", "ish", "guo"))),
            by = "imo") |> 
  mutate(
    ish_confidence = case_when(
      !is.na(ish_name) ~ "1: Direct IMO ORBIS match",
      TRUE ~ "Other"
    )
  )

matches_1 <- ownership_table_1 |> filter(ish_confidence == "1: Direct IMO ORBIS match")

# For the rest, we try to complete using ORBIS result for the search based on GISIS company name 
ownership_table_2 <- ownership_table_1 |> 
  filter(ish_confidence == "Other") |> 
  dplyr::select(vessel_id, mmsi, vessel_name, imo_confidence_code, imo_confidence_status, imo) |> 
  left_join(gfw_gisis_key |> 
              dplyr::select(imo, company_imo) |> 
              filter(!is.na(imo) & !is.na(company_imo)) |> 
              distinct(),
            by = "imo") |> 
  left_join(orbis_companies |> 
              dplyr::select(starts_with(c("gisis", "ish", "guo"))),
            by = join_by(company_imo == gisis_company_imo_number),
            na_matches = "never") |> 
  dplyr::select(-c(company_imo, gisis_owner_name)) |> 
  mutate(
    ish_confidence = case_when(
      !is.na(ish_name) ~ "2: GISIS owner ORBIS match",
      TRUE ~ "Other"
    )
  )
  
matches_2 <- ownership_table_2 |> filter(ish_confidence == "2: GISIS owner ORBIS match")

# For the vessels we could not match in ORBIS, we use the GISIS owner as ISH
ownership_table_3 <- ownership_table_2 |> 
  filter(ish_confidence == "Other") |> 
  dplyr::select(vessel_id, mmsi, vessel_name, imo_confidence_code, imo_confidence_status, imo) |> 
  left_join(gfw_gisis_key |> 
              distinct(imo, registered_owner, company_nationality) |> 
              filter(!grepl("Rptd", registered_owner)),                    # When GISIS mentions an unknown owner, we encode it as NA instead of adding poor quality info on nationality
            by = "imo") |> 
  mutate(
    company_nationality = case_when(
      company_nationality == "Azores" ~ "Portugal",
      company_nationality == "Canary Islands" ~ "Spain",
    ),
    company_nationality = countrycode(company_nationality, origin = "country.name", destination = "iso2c")
  ) |> 
  rename(ish_name = registered_owner, ish_country_iso2 = company_nationality) |> 
  mutate(
    ish_confidence = case_when(
      !is.na(ish_name) ~ "3: GISIS owner as ISH",
      TRUE ~ "4: No known ISH"
    )
  )
  

# Combine all the vessels

ownership <- bind_rows(matches_1, matches_2, ownership_table_3) |> 
  mutate(
    ish_country_iso2 = countrycode(ish_country_iso2, origin = "iso2c", destination = "iso3c"),
    guo_country_iso2 = countrycode(guo_country_iso2, origin = "iso2c", destination = "iso3c")
  ) |> 
  rename(
    ish_country_iso3 = ish_country_iso2,
    guo_country_iso3 = guo_country_iso2
  )

# Lastly, for the vessels with an ISH but no GUO, we use the ISH as GUO

# Flag the GUO confidence
ownership <- ownership |> 
  mutate(
    guo_confidence = case_when(
      !is.na(guo_name) ~ "1: GUO found by ORBIS",
      !is.na(ish_name) ~ "2: ISH used as GUO",
      TRUE ~ "3: No known GUO"
    )
  )

# Complete GUO with ISH
ownership <- ownership |> 
  mutate(
    guo_name = case_when(is.na(guo_name) ~ ish_name, TRUE ~ guo_name),
    guo_country_iso3 = case_when(is.na(guo_country_iso3) ~ ish_country_iso3, TRUE ~ guo_country_iso3),
    guo_type = case_when(is.na(guo_type) ~ ish_type, TRUE ~ guo_type)
  )

# Add total confidence level
ownership <- ownership |> 
  mutate(
    total_confidence = case_when(
      ish_confidence == "1: Direct IMO ORBIS match" & guo_confidence == "1: GUO found by ORBIS" ~ "High",
      ish_confidence == "2: GISIS owner ORBIS match" & guo_confidence == "1: GUO found by ORBIS" ~ "High",
      ish_confidence == "2: GISIS owner ORBIS match" & guo_confidence == "2: ISH used as GUO" ~ "Medium",
      ish_confidence == "3: GISIS owner as ISH" & guo_confidence == "2: ISH used as GUO" ~ "Low",
      TRUE ~ NA
    ),
    total_confidence = factor(total_confidence, levels = c("Low", "Medium", "High"))
  )
```

Two pieces of information about ownership obtained separately are added: *Altantic Shrimpers Ltd* is owned by the Dutch *Cornelis Vrolijk* and *HISEPEC* is owned by the Spanish *PEIRERA GROUP*.

```{r}
# Adding info about Atlantic Shrimpers and HISEPEC
mapping <- tibble(
  guo_name_pattern = c("ATLANTIC SHRIMPERS LTD", "HISEPEC"),
  new_guo_name = c("CORNELIS VROLIJK HOLDING B.V.", "PEREIRA GROUP"),
  new_iso3 = c("NLD", "ESP"),
)

ownership <- ownership |> 
  rowwise() |> 
  mutate(
    matched = mapping |> 
      filter(str_detect(guo_name, guo_name_pattern)) |> 
      slice(1) |> # In case of multiple matches, take the first
      list(),
    guo_name = ifelse(nrow(matched) > 0, matched[1,]$new_guo_name, guo_name),
    guo_country_iso3 = ifelse(nrow(matched) > 0, matched[1,]$new_iso3, guo_country_iso3)
  ) |> 
  dplyr::select(-matched) |> 
  ungroup()
```

Catching an error in the database:

```{r}
ownership <- ownership |> 
  mutate(imo = case_when(
    mmsi == 657216600 ~ 8778005,   # fixing a mistake
    TRUE ~ imo
  ))
```

Rename the IMO number variable for clarity:

```{r}
ownership <- ownership |> 
  rename(assigned_imo = imo)
```

Save for later use:

```{r}
write.xlsx(ownership,
           "output/data/ownership/ownership_data_FAO34_2025-04_GFW_effort_2020-2024.xlsx",
           as.table = TRUE)
```

```{r}
rm(list = c("effort", "vessels"))
rm(list = ls()[grepl("imo|table|orbis|matches|owners_list|mapping", ls())])
```

### Joining with apparent fishing effort data

```{r}
join_ownership <- function(gfw_ds,
                           ownership_data,
                           key = c("vessel_id", "mmsi"),
                           keep_ambiguous_imos = FALSE) {
  
  key <- match.arg(key)
  
  if (!keep_ambiguous_imos) {
    ownership_data <- ownership_data |> 
      mutate(across(
        ish_name:total_confidence,
        ~ ifelse(imo_confidence_code == 4, NA, .)
      ))
  }
  
  gfw_ds$clean_effort <- gfw_ds$clean_effort |> 
    left_join(ownership_data,
              by = key,
              relationship = "many-to-one")
  
  if (!is.null(gfw_ds$small_efforts)) {
    gfw_ds$small_efforts <- gfw_ds$small_efforts |> 
      left_join(ownership_data,
                by = key,
                relationship = "many-to-one")
  }
  
  return(gfw_ds)
}

gfw_ds_ML <- readRDS("output/clean_data/_temp/gfw_effort_FAO34_ML_0.rds")
gfw_ds_YH <- readRDS("output/clean_data/_temp/gfw_effort_FAO34_YH_0.rds")

ownership_by_vessel_id <- ownership |> 
  dplyr::select(-c(mmsi, vessel_name)) |> 
  distinct()

gfw_ds_ML <- join_ownership(gfw_ds = gfw_ds_ML, 
                            ownership_data = ownership_by_vessel_id,
                            key = "vessel_id")
gfw_ds_YH <- join_ownership(gfw_ds = gfw_ds_YH,
                            ownership_data = ownership_by_vessel_id,
                            key = "vessel_id")

# MONTHLY LOW is save as a final dataset
if (!file.exists("output/clean_data/apparent_fishing_effort.rds") | FORCE_WRITE) {
  saveRDS(gfw_ds_ML, "output/clean_data/apparent_fishing_effort.rds")
}
# YEARLY HIGH is save as temporary since it will be further transformed
if (!file.exists("output/clean_data/_temp/gfw_effort_FAO34_YH_1.rds") | FORCE_WRITE) {
  saveRDS(gfw_ds_YH, "output/clean_data/_temp/gfw_effort_FAO34_YH_1.rds")
}
```

```{r}
rm(list = setdiff(ls()[grepl("gfw", ls())], lsf.str()))
gc()
```

## Reconstructed catches and stock assessments

### Reconstructed catches from Sea Around Us

We compile the reconstructed catch datasets from the [Sea Around Us](https://www.seaaroundus.org) project. These data correspond to the Sea Around Us spatialized catches contained within three polygons: FAO area 34, the Canary Current LME (LME 27), and the Guinea Current LME (LME 28).

Since the LMEs are almost entirely inscribed in the FAO area, we deduct the LME catches from those of the FAO area. This gives the catches on the high seas, as opposed to the catches in the LMEs, which cover most of the EEZs and the entire continental shelf of the region.

Since the LMEs are not entirely inscribed in the FAO zone (in particular LME 28), negative values are generated when computing the high seas catches. In those cases the high seas catch is set to 0 t for the taxon.

```{r}

# Helper function (the csv downloaded from the website contained missing ',' when the last column in NA)

fix_csv_trailing_missing_field <- function(input_csv, output_csv, expected_cols = 15) {
  lines <- read_lines(input_csv)
  
  # Store header and data separately
  header <- lines[1]
  data_lines <- lines[-1]
  
  # Define expected number of columns from header
  expected_cols <- str_count(header, ",") + 1
  
  fixed_data_lines <- vapply(data_lines, function(line) {
    # Split on commas outside quotes
    fields <- strsplit(line, ",(?=(?:[^\"]*\"[^\"]*\")*[^\"]*$)", perl = TRUE)[[1]]
    
    if (length(fields) == expected_cols - 1) {
      # Insert empty field as last column (landed_value missing)
      fields <- c(fields, "")
    }
    
    paste(fields, collapse = ",")
  }, FUN.VALUE = character(1))
  
  # Write back with original header
  write_lines(c(header, fixed_data_lines), output_csv)
  
  message("CSV fixed and saved to: ", output_csv)
  
  invisible(TRUE)
}

# Fix missing comma issue

files <- c(
  "data/catch/reconstructed_catch/SAU FAO 34 v50-1/SAU FAO 34 v50-1",
  "data/catch/reconstructed_catch/SAU LME 27 v50-1/SAU LME 27 v50-1",
  "data/catch/reconstructed_catch/SAU LME 28 v50-1/SAU LME 28 v50-1"
  )

if (!file.exists("data/catch/reconstructed_catch/SAU FAO 34 v50-1/SAU FAO 34 v50-1_fixed.csv") | FORCE_WRITE) {
  sapply(files, 
       function(f) {
         fix_csv_trailing_missing_field(
           input_csv = paste0(f, ".csv"), 
           output_csv = paste0(f, "_fixed.csv")
         )
       }
  )
}


# Load fixed data

catch_tot <- read_csv("data/catch/reconstructed_catch/SAU FAO 34 v50-1/SAU FAO 34 v50-1_fixed.csv") |> clean_names()
catch_lme27 <- read_csv("data/catch/reconstructed_catch/SAU LME 27 v50-1/SAU LME 27 v50-1_fixed.csv") |> clean_names()
catch_lme28 <- read_csv("data/catch/reconstructed_catch/SAU LME 28 v50-1/SAU LME 28 v50-1_fixed.csv") |> clean_names()

catch_lme <- catch_lme27 |> 
  mutate(area_name = "LME 27") |> 
  rbind(catch_lme28 |> 
          mutate(area_name = "LME 28"))

cols <- colnames(catch_lme)
n <- length(cols)
cols <- cols[-c(1, n - 1, n)]

catch_lme_tot <- catch_lme |> 
  summarize(
    tonnes = sum(tonnes, na.rm = TRUE),
    landed_value = sum(landed_value, na.rm = TRUE),
    .by = all_of(cols)
  )

cols <- cols[-1]

catch_high_seas <- catch_tot |>
  left_join(
    catch_lme_tot |> dplyr::select(-area_type),
    by = cols
  ) |> 
  mutate(
    tonnes = pmax(tonnes.x - tonnes.y, 0),                      # In some cases the catch is higher in the LMEs than in the whole FAO 34.
    landed_value = pmax(landed_value.x - landed_value.y, 0)     # Prob. bc LME 28 isn't fully contained in FAO 34.
  ) |> 
  dplyr::select(-c(tonnes.x, tonnes.y, landed_value.x, landed_value.y)) |> 
  mutate(area_name = "FAO 34 - {LMEs}")

catch <- rbind(
  catch_lme,
  catch_high_seas
)

rm(list = c("catch_high_seas", "catch_lme", "catch_lme_tot", "catch_lme27", "catch_lme28", "catch_tot", "cols", "files", "n"))
```

Only the catch of the "Industrial" sector during the 2015-2019 period is kept:

```{r}
catch <- catch |> 
  filter(fishing_sector == "Industrial") |> 
  filter(year >= 2015, year <= 2019)
```

### Joining with CECAF and ICCAT stock assessments

Assessment data from reports by the FAO-CECAF expert group and the International Commission for the Conservation of Atlantic Tunas (ICCAT) are loaded. The CECAF data comes from the reports of the northern and southern sub-groups, for small pelagics on the one hand and demersal species on the other. When several stocks are contained within the same sub-group region, they are aggregated into a single stock. The assessment status for this “super-stock” is inherited from the actual stocks: if all stocks have the same status, this status is retained; otherwise, the status associated with the stock with the highest catches reported to the FAO in the year of the assessment is retained. In the latter case, the status is preceded by the word “Mainly.”

```{r}
cecaf <- read.xlsx("data/iccat_cecaf_stock_evaluations.xlsx", sheet = "cecaf_clean")  |> clean_names()
iccat <- read.xlsx("data/iccat_cecaf_stock_evaluations.xlsx", sheet = "iccat_clean") |> clean_names()
```

The assessment data is joined to the catch data. Since the CECAF sub-group regions correspond almost perfectly to LMEs 27 (northern subgroup) and 28 (southern subgroup), the stocks fished in an LME are allocated the assessments of the corresponding subgroup. The scientific name of the species is used as the join key within the LME-subgroup subset.

In the case of CECAF assessments, a stock may not be defined at the specific level, but rather at the genus level (e.g., *Sepia* spp.) or family level (e.g., *Sparidae*). In some cases, one stock is defined for a species (e.g., *Penaeus notialis* in the southern subgroup) and another is defined for the genus of that species in the same subregion (*Penaeus* spp. in the south).

To assign them CECAF assessments, we proceed as follows. For each LME-CECAF subgroup pair:

1.  We perform an initial join on the full scientific name (exact_match).
2.  The remaining catches are joined on the first part of the name (higher_level_match). In addition to catches defined at the genus level, catches defined at the species level but not associated in exact_match can be associated.
3.  These tables are then stacked.

For ICCAT assessments, stocks are always defined at the species level. Joins are made only on the full scientific name.

Sea Around Us catches are also defined at varying taxonomic levels (cf. next paragraph).

```{r}

match_cecaf_assessment <- function(sau_catch, cecaf){
  
  # Prepare catch data by separating the taxonomic levels
  
  catch <- catch |> 
    separate(
      scientific_name,
      into = c("catch_taxa_higher", "catch_taxa_lower"),
      sep = " ",
      remove = FALSE
    )

  # Match to CECAF assessments (although only valid for recent years)
  
  cecaf_mod <- cecaf |> 
    dplyr::select(species, cecaf_subregion, assessment_fr, assessment_level) |> 
    separate(
      species,
      into = c("eval_taxa_higher", "eval_taxa_lower"),
      sep = " ",
      remove = FALSE
    ) |> 
    mutate(
      eval_lme = case_when(
        cecaf_subregion == "North-Western Africa" ~ "LME 27",
        cecaf_subregion == "South" ~ "LME 28"
      )
    ) |> 
    rename(
      cecaf_assessment_fr = assessment_fr, 
      cecaf_assessment_level = assessment_level
    )
  
  # Join by exact match
  exact_match <- catch |> 
    left_join(
      cecaf_mod,
      by = c("catch_taxa_higher" = "eval_taxa_higher", "catch_taxa_lower" = "eval_taxa_lower"),
      relationship = "many-to-many"                             # many-to-many bc. northern and southern stocks are present in evaluation
    ) |> 
    filter(area_name == eval_lme)                               # here we filter so that eventually we have a many-to-one
  
  # Join the remainder by higher-level match (catch at species, CECAF at genus level)
  no_exact_match <- setdiff(
    catch,
    exact_match[, colnames(catch)]
  )
  
  higher_level_match <- no_exact_match |> 
    left_join(
      cecaf_mod |> filter(is.na(eval_taxa_lower) | eval_taxa_lower %in% c("sp.", "spp.")),
      by = c("catch_taxa_higher" = "eval_taxa_higher"),
      relationship = "many-to-many"                            
    ) |> 
    filter(area_name == eval_lme)
  
  # Combine both
  cecaf_match <- bind_rows(exact_match, higher_level_match)
  
  no_cecaf_match <- setdiff(
    catch,
    cecaf_match[, colnames(catch)]
  )
  
  catch <- bind_rows(cecaf_match, no_cecaf_match) |> 
    dplyr::select(-c(species, eval_taxa_lower, cecaf_subregion, eval_lme, catch_taxa_higher, catch_taxa_lower))
  
  # Return result
  return(catch)
}


match_iccat_assessment <- function(sau_catch, iccat) {
  
  # Prepare catch data by separating the taxonomic levels
  
  catch <- catch |> 
    separate(
      scientific_name,
      into = c("catch_taxa_higher", "catch_taxa_lower"),
      sep = " ",
      remove = FALSE
    )
  
  # Match to ICCAT assessments (although only valid for recent year)
  
  iccat_mod <- iccat |> 
    dplyr::select(species, overfishing, overfished) |> 
    separate(
      species,
      into = c("eval_taxa_higher", "eval_taxa_lower"),
      sep = " ",
      remove = FALSE
    )
  
  # Join by exact match
  catch <- catch |> 
    left_join(
      iccat_mod,
      by = c("catch_taxa_higher" = "eval_taxa_higher", "catch_taxa_lower" = "eval_taxa_lower"),
      relationship = "many-to-one"
    ) |> 
    dplyr::select(-c(species, catch_taxa_higher, catch_taxa_lower))
  
  # No higher level match bc. all assessments are at the species level
  
  return(catch)

}

catch <- match_cecaf_assessment(catch, cecaf)
catch <- match_iccat_assessment(catch, iccat)
```

Catches for which the taxonomic level is less precise than the one used by the expert groups are not matched. They are labeled as “Match impossible” to distinguish between uncertainty related to reconstruction and the absence of evaluation.

-   Taxa containing “not identified” in their name are directly categorized as impossible matches.

-   For catches of “single-word” taxa, i.e., identified at a level higher than the species, the category “Match impossible” is assigned if the captured taxon is contained in the lineage of an assessed taxon (i.e. higher taxonomic resolution of the assessment), and “Non évalué" otherwise.

```{r}

if (!SKIP_CATCH) {
  clean_taxon_name <- function(name) {
    str_replace(name, "^([A-Za-z]+)\\s+(sp\\.|spp\\.)$", "\\1")
  }
  
  get_lineage <- function(name) {
    
    name <- clean_taxon_name(name)
    
    # Get AphiaID
    id <- tryCatch(wm_records_name(name)$AphiaID[1], error = function(e) NA)
  
    # If either ID is missing, return NA
    if (is.na(id)) return(NA)
  
    # Get classification of B
    lineage <- tryCatch(wm_classification(id), error = function(e) NA)
    if (is.data.frame(lineage)) {
      return(paste(lineage$scientificname, collapse = "/"))
    } else {
      return(NA)
    }
  }
  
  # Immediate impossible matches : taxa not identified
  
  immediate_impossible_matches <- catch |> 
    filter(grepl("not identified", scientific_name)) |> 
    mutate(impossible_match = TRUE) |> 
    distinct(scientific_name, area_name, impossible_match)
  
  # Potential : taxa identified but not matched, possibly because of coarse taxonomic resolution
  
  potential_impossible_matches <- catch |>
    filter(!grepl("not identified", scientific_name)) |> 
    filter(is.na(cecaf_assessment_fr) & is.na(overfishing) & is.na(overfished)) |> 
    distinct(scientific_name, area_name)
  
  # Extract lineages from assessment scientific names
  
  cecaf_w_lineage <- cecaf |>
    filter(!is.na(species)) |> 
    rowwise() |> 
    mutate(lineage = get_lineage(species)) |> 
    ungroup()
  
  iccat_w_lineage <- iccat |> 
    rowwise() |> 
    mutate(lineage = get_lineage(species)) |> 
    ungroup()
  
  # Find impossible matches for CECAF species (by LME-subgroup)
  
  impossible_matches_cecaf_north <- potential_impossible_matches |> 
    filter(area_name == "LME 28") |> 
    cross_join(cecaf_w_lineage |> 
                 filter(cecaf_subregion == "North-Western Africa") |>
                 dplyr::select(lineage)) |> 
    rowwise() |> 
    mutate(
      impossible_match = grepl(scientific_name, lineage)
    ) |> 
    ungroup() |> 
    filter(impossible_match) |> 
    distinct(scientific_name, area_name, impossible_match)
  
  impossible_matches_cecaf_south <- potential_impossible_matches |> 
    filter(area_name == "LME 27") |> 
    cross_join(cecaf_w_lineage |> 
                 filter(cecaf_subregion == "South") |>
                 dplyr::select(lineage)) |> 
    rowwise() |> 
    mutate(
      impossible_match = grepl(scientific_name, lineage)
    ) |> 
    ungroup() |> 
    filter(impossible_match) |> 
    distinct(scientific_name, area_name, impossible_match)
  
  # Find impossible matches for ICCAT species
  
  impossible_matches_iccat <- potential_impossible_matches |> 
    cross_join(iccat_w_lineage |> 
                 dplyr::select(lineage)) |> 
    rowwise() |> 
    mutate(
      impossible_match = grepl(scientific_name, lineage) 
    ) |> 
    ungroup() |> 
    filter(impossible_match) |> 
    distinct(scientific_name, area_name, impossible_match)
  
  # Bind the impossible matches
  
  impossible_matches <- bind_rows(
    immediate_impossible_matches,
    impossible_matches_cecaf_north,
    impossible_matches_cecaf_south,
    impossible_matches_iccat
  ) |> 
    distinct() # Some high taxonomic levels might be shared between ICCAT and CECAF species
  
  catch <- catch |> 
    left_join(
      impossible_matches,
      by = c("scientific_name", "area_name"),
      relationship == "many-to-one"
    ) |> 
    mutate(
      impossible_match = ifelse(is.na(impossible_match), FALSE, TRUE)
    )
}
```

Eventually the official assessments are harmonized with a unique label variable:

```{r}

if (!SKIP_CATCH) {
  catch <- catch |> 
    mutate(
      official_status = case_when(
        !is.na(cecaf_assessment_fr) ~ str_trim(cecaf_assessment_fr),
        overfished == 1 & overfishing == 1 ~ "Surpêché et dégradé",
        overfished == 1 ~ "Dégradé",
        overfishing == 1 ~ "Surpêché",
        overfished == 0 & overfishing == 0 ~ "Ni dégradé ni surpêché",
        impossible_match ~ "Match impossible",
        TRUE ~ "Non évalué"
      ),
      dim_assessment = case_when(
        !is.na(cecaf_assessment_fr) ~ 1,
        !is.na(overfished) | !is.na(overfishing) ~ 2,
        TRUE ~ 0
      ),
      official_status = factor(
        official_status,
        levels = c("Ni dégradé ni surpêché",
                   "Non pleinement exploité",
                   "Principalement non pleinement exploité",
                   "Pleinement exploité",
                   "Principalement pleinement exploité",
                   "Principalement surexploité",
                   "Surpêché",
                   "Dégradé",
                   "Surpêché et dégradé",
                   "Surexploité",
                   "Principalement non évalué",
                   "Non évalué",
                   "Match impossible")
      )
    )
}
```

The final overexploitaiton label is edited to distinguish between stocks with at least one overpoitation indicator ("surexploité" or "surpêché" or "dégradé"), from stocks having none. This final label contains the mention "probable" to underline its aggregated and uncertain nature. "Non évalué" and "Match impossible" values are kept as is.

```{r}

if (!SKIP_CATCH) {
  catch <- catch |> 
    mutate(
      overexploitation_flag = case_when(
        official_status %in% c("Surexploité", "Principalement surexploité", "Dégradé", "Surpêché", "Surpêché et dégradé") ~ "Surexploitation probable",
        official_status %in% c("Ni dégradé ni surpêché", "Non pleinement exploité", "Principalement non pleinement exploité", "Pleinement exploité", "Principalement pleinement exploité") ~  "Non surexploitation probable",
        official_status %in% c("Non évalué", "Principalement non évalué") ~  "Evaluations lacunaires",
        official_status == "Match impossible" ~ "Match impossible"
      )
    )
}
```

### Saving

```{r}
if (!SKIP_CATCH) {saveRDS(catch, "output/clean_data/catch.rds")}
```

```{r}
rm(list = ls()[grepl("impossible|cecaf|iccat|catch", ls())])
gc()
```

## Global Fishing Watch apparent trawling effort

We load unpublished data provided by Global Fishing Watch. This is the fishing effort of the global trawler fleet visible by AIS from 2020 to 2023. Regardless of the vessel, each fishing time observation is categorized as “bottom trawling,” “pelagic trawling,” or “other.” The spatial resolution is 0.1°x0.1° and the temporal resolution is one month.

```{r}
#| include: false
trawling_files <- c(list.files(path = "data/raw_gfw_trawling_effort/12424720-19d4-11f0-beb1-9d682effc649", pattern = "\\.csv$", full.names = TRUE),
                   list.files(path = "data/raw_gfw_trawling_effort/f18b56c0-19d3-11f0-9018-5df6c30331c3", pattern = "\\.csv$", full.names = TRUE))

raw_bottom_trawling <- lapply(trawling_files, read_csv) |> 
  bind_rows()
```

### First cleaning

Due to the poorer performance of the model in the “pelagic trawling” class, only bottom trawling observations are retained.

```{r}
bottom_trawling <- raw_bottom_trawling |> 
  filter(label == "bottom_trawling")
```

A spatial filter is applied to the data in order to retain only the effort made in FAO area 34.

```{r}
# First filter based on bounding box (faster)
bbox <- st_bbox(fao34_sf)

bottom_trawling <- bottom_trawling |> 
  filter(
    cell_ll_lon >= bbox["xmin"] & cell_ll_lon <= bbox["xmax"],
    cell_ll_lat >= bbox["ymin"] & cell_ll_lon <= bbox["ymax"]
  )

bottom_trawling <- bottom_trawling |> 
  st_as_sf(coords = c("cell_ll_lon","cell_ll_lat"), remove = FALSE, crs = st_crs(4326))

# Then based on the FAO 34 limits
bottom_trawling <- bottom_trawling |> 
  st_filter(fao34_sf) |> 
  st_drop_geometry()
```

The columns are renamed to obtain data similar to the apparent fishing effort data processed previously.

```{r}
bottom_trawling <- bottom_trawling |> 
  rename(
    lat = cell_ll_lat, 
    lon = cell_ll_lon,
    apparent_fishing_hours = fishing_hours,
  ) |> 
  mutate(
    time_range = paste(year, month, sep = "-")
  ) |>
  dplyr::select(-c(month, n_mmsi, n_events)) |> 
  relocate(
    lat, 
    lon,
    year,
    time_range,
    year,
    flag,
    mmsi,
    label,
    apparent_fishing_hours,
    distance_trawled_km)
```

Cleaning is similar to that of Global Fishing Watch apparent fishing effort data. It is slightly simpler since trawling data does not contain a `gear_type` attribute, IMO number, vessel name, or call sign.

```{r}
bottom_trawling <- bottom_trawling |> 
  spatial_join_gfw_data() |> 
  add_political_labels() |> 
  add_vessel_characteristics() |>
  mutate(kWh = engine_power_kw_gfw * apparent_fishing_hours) |> 
  dplyr::select(-label) |> 
  relocate(
    year, time_range,
    lat, lon,
    region_type, region_type_fr,
    eez_name, eez_iso3c,
    flag, flag_iso3c,
    mmsi,
    apparent_fishing_hours, distance_trawled_km, kWh,
    #is_european_flag, 
    is_fao34_flag, is_west_african_flag,
    domestic, origin_type_fr, destination_type_fr,
    length_m_gfw, engine_power_kw_gfw, tonnage_gt_gfw,
    distance_to_shore_km, distance_from_port_km, bathymetry_m,
  )
```

We keep the effort that meets the `min_dist` and `min_hours` thresholds, as with the apparent fishing effort from Global Fishing Watch.

```{r}
bottom_trawling_ds <- clean_gfw_data(
  bottom_trawling,
  min_dist = 5,
  min_hours = 2000,
  unit = "mmsi",
  save_small_efforts = TRUE
)
```

### Joining with ownership data

```{r}
ownership_by_mmsi <- ownership |> 
  dplyr::select(-c(vessel_id, vessel_name)) |> 
  filter(imo_confidence_code != 4) |> # Remove IMOs with poor confidence
  dplyr::select(-contains("confidence")) |> 
  distinct()

bottom_trawling_ds <- join_ownership(
  gfw_ds = bottom_trawling_ds,
  ownership_data = ownership_by_mmsi,
  key = "mmsi",
  keep_ambiguous_imos = TRUE # Poor confidence IMOs have already been removed
)
```

### Bottom gear width

We draw on the work of Eigaard et al. (2016), which models the opening width of bottom trawls based on vessel characteristics.

The models developed by Eigaard et al. are adjusted using “BENTHIS métiers” These categories are created by combining the gear (e.g., otter trawl, beam trawl, demersal seine) and the categories of target species (e.g., crustaceans, demersal fish).

```{r}

eigaard <- read_csv("data/eigaard/BENTHIS_otter_trawl_data.csv") |> clean_names()

# library(ggplot2)
# 
# eigaard |> 
#   filter(!is.na(door_spread)) |> 
#   ggplot(aes(x = door_spread)) +
#   geom_histogram()
```

In the absence of precise information, two assumptions are made. First, it is assumed that all bottom trawls used in FAO zone 34 are otter trawls. In addition, all trawlers are assigned the OT_MIX gear type. This is the broadest category in terms of target species.

We know that the fleet belonging to Atlantic Shrimpers Ltd consists exclusively of shrimp trawlers. It would be possible to adapt the BENTHIS métier for this fleet, choosing OT_CRU, OT_MIX_CRU, or OT_MIX_CRU_DMF instead. Without being able to choose between these categories, we retain OT_MIX for all trawlers for now.

We can then calculate the area covered by bottom trawling observations.

To estimate a conservative spatial footprint for bottom trawling, we also assign a conservative opening between the doors of 25 m.

```{r}

add_bottom_impact <- function(bottom_trawling_data) {
  
  bottom_trawling_data |> 
    filter(!is.na(engine_power_kw_gfw)) |> 
    mutate(
      gear_width_model = "OTB_MIX",
      model_a = 10.6608,
      model_b =  0.2921,
      std_error_a = 6.6939,
      std_error_b = 0.1044,
      gear_width_m = model_a * (engine_power_kw_gfw ^ model_b),
      gear_width_minimal_m = 25
    ) |> 
    mutate(
      swept_area_km2 = gear_width_m * 1e-3 * distance_trawled_km
    )

}

bottom_trawling_ds$clean_effort <- bottom_trawling_ds$clean_effort |> add_bottom_impact()
bottom_trawling_ds$small_efforts <- bottom_trawling_ds$small_efforts |> add_bottom_impact()
```

In order to compare the area trawled in a cell with the area of the cell, the latter is calculated. This is because the area of a cell with dimensions defined in degrees varies depending on factors such as latitude. The `geosphere` package is used for this purpose.

```{r}
library(geosphere)

compute_cell_area <- function(lat, lon, height_degree = 0.1, width_degree = 0.1) {

  dlat <- height_degree / 2
  dlon <- width_degree / 2
  
  # Define corners
  p1 <- c(lon - dlon, lat - dlat)  # lower left
  p2 <- c(lon + dlon, lat - dlat)  # lower right
  p3 <- c(lon + dlon, lat + dlat)  # upper right
  p4 <- c(lon - dlon, lat + dlat)  # upper left
  
  # Build polygon
  cell <- rbind(p1, p2, p3, p4, p1)
  
  # Compute area in m2 and convert to km2
  area_km2 <- geosphere::areaPolygon(cell) / 1e6
  
  return(area_km2)
}

bottom_trawling_ds$clean_effort <- bottom_trawling_ds$clean_effort |> 
  rowwise() |> 
  mutate(cell_area_km2 = compute_cell_area(lat, lon)) |> 
  ungroup()

bottom_trawling_ds$small_efforts <- bottom_trawling_ds$small_efforts |> 
  rowwise() |> 
  mutate(cell_area_km2 = compute_cell_area(lat, lon)) |> 
  ungroup()
```

### Laying a grid above the continental shelf

In order to calculate spatial footprint statistics - particularly surface area proportions - a complete grid is created above the continental shelf (approximated as regions with bathymetry \>= 1000 m).

```{r}
#| include: false

# Creating the grid
bbox <- st_bbox(fao34_sf)

lats <- seq(ceiling(bbox["ymin"] * 10) / 10, floor(bbox["ymax"] * 10) / 10, by = 0.1)
lons <- seq(ceiling(bbox["xmin"] * 10) / 10, floor(bbox["xmax"] * 10) / 10, by = 0.1)

grid_centers <- expand.grid(lon = lons, lat = lats) |> 
  mutate(lat = round(lat, 1), lon = round(lon, 1))

grid <- st_as_sf(grid_centers, coords = c("lon", "lat"), crs = st_crs(fao34_sf), remove = FALSE)
grid <- grid[st_within(grid, fao34_sf, sparse = FALSE)[,1], ]

# Extracting bathymetry
grid$bathymetry_m <- raster::extract(bathymetry_raster, grid)

# Adding EEZ name etc
grid <- grid |> spatial_join_gfw_data()

# Filtering grid to keep shelf
shelf_grid <- grid |> 
  filter(
    bathymetry_m >= -1000, 
    bathymetry_m <= 0,
    distance_to_shore_km >= 1,
    !is.na(eez_iso3c)
  )

# Adding cell area
shelf_grid <- shelf_grid |> 
  st_drop_geometry() |> 
  rowwise() |> 
  mutate(cell_area_km2 = compute_cell_area(lat, lon)) |> 
  ungroup()

# Add to bottom trawling dataset
bottom_trawling_ds$shelf_grid <- shelf_grid
```

### Saving

```{r}
saveRDS(bottom_trawling_ds, "output/clean_data/bottom_trawling.rds")
```

```{r}
#| include: false

rm(list = ls()[grepl("bottom|ownership|vessel", ls())])
gc()
```

## Fishing close to shore

Labels characterizing fishing in coastal areas (12-mile bands or restricted areas) are added. These labels are added to the high spatial resolution fishing effort dataset.

```{r}

inshore_fishing_ds <- readRDS("output/clean_data/_temp/gfw_effort_FAO34_YH_1.rds")

# Restricted inshore areas
restricted <- read.xlsx("data/inshore_restricted_areas_FAO34.xlsx", sheet = "clean") |> 
  mutate(
    total_km = 1.852 * total_nm,
    partial_km = 1.852 * partial_nm
  ) |> 
  filter(!(eez_iso3c %in% c("ESP", "PRT", "SHN", "BRA"))) |>      # Keep only west african countries
  mutate(
    total_km = ifelse(is.na(total_km), 0, total_km),
    partial_km = ifelse(is.na(partial_km), 0, partial_km),
    total_nm = ifelse(is.na(total_nm), 0, total_nm),
    partial_nm = ifelse(is.na(partial_nm), 0, partial_nm),
  )

# Saving the restricted zones in the dataset
inshore_fishing_ds$restricted_areas <- restricted

# Flagging fishing observations
inshore_fishing_ds$clean_effort <- inshore_fishing_ds$clean_effort |> 
  left_join(restricted,
            by = "eez_iso3c") |> 
  mutate(
    inn_fishing = case_when(
      distance_to_shore_km < total_km ~ TRUE,
      distance_to_shore_km < partial_km & !is.na(prohibited_flag) & !is.na(prohibited_gear_type) & !domestic & gear_type == "Trawls" ~ TRUE,
      distance_to_shore_km < partial_km & !is.na(prohibited_flag) & !domestic ~ TRUE,
      TRUE ~ FALSE
    ),
    domestic_guo = (guo_country_iso3 == eez_iso3c) | (eez_iso3c == "ESH") & (guo_country_iso3 == "MAR"),
    would_be_inn = case_when(
      is.na(domestic_guo) ~ NA,
      distance_to_shore_km < total_km ~ TRUE,
      distance_to_shore_km < partial_km & !is.na(prohibited_flag) & !is.na(prohibited_gear_type) & !domestic_guo & gear_type == "Trawls" ~ TRUE,
      distance_to_shore_km < partial_km & !is.na(prohibited_flag) & !domestic_guo ~ TRUE,
      TRUE ~ FALSE
    ),
    in_12nm = (distance_to_shore_km < 12 * 1.852)
  )
```

We create a subset that excludes international waters and contains only trawler effort, which is less cumbersome to load.

```{r}

inshore_fishing_shelf <- inshore_fishing_ds$clean_effort |> 
  filter(!is.na(eez_iso3c), gear_type == "Trawls")
```

### Saving

```{r}
saveRDS(inshore_fishing_ds, "output/clean_data/inshore_fishing.rds")
saveRDS(inshore_fishing_shelf, "output/clean_data/inshore_fishing_eez_trawlers.rds")
```
